A few stickleback suffice for the transport of alleles to new lakes

### 2019-09-26 Stickleback_Paper.tex
Bill's last edits

Threespine stickleback populations provide a striking example of local adaptation to divergent habitats 
in populations that are connected by recurrent gene flow. 
These small fish occur in marine and freshwater habitats throughout the Northern Hemisphere, and in numerous cases the smaller freshwater populations have been established ``de novo'' from marine colonists. 
Independently evolved freshwater populations show similar phenotypes that have been shown to derive largely from the same standing genetic variants.
Geographic isolation prevents direct migration between the freshwater populations, strongly suggesting that these shared locally adaptive alleles are transported through the marine population.
However it is still largely unknown how gene flow, recombination, and selection jointly impact the standing variation that might fuel this adaptation.
Here we use individual-based, spatially explicit simulations to determine the levels of gene flow that best match observed patterns of allele sharing among habitats in stickleback.
We aim to better understand how gene flow and local adaptation in large metapopulations determine the speed of adaptation and re-use of standing genetic variation. 
In our simulations we find that repeated adaptation uses a shared set of alleles that are maintained at low frequency by migration-selection balance in oceanic populations. 
This process occurs over a realistic range of intermediate levels of gene flow that match previous empirical population genomic studies in stickleback. 
Examining these simulations more deeply reveals how lower levels of gene flow leads to slow, independent adaptation to different habitats, whereas higher levels gene flow leads to significant mutation load -- but an increased probability of successful population genomic scans for locally adapted alleles. 
Surprisingly, we find that the genealogical origins of most freshwater adapted alleles can be traced back to the original generation of marine individuals that colonized the lakes, as opposed to subsequent migrants. 
These simulations provide deeper context for existing studies of stickleback evolutionary genomics, and guidance for future empirical studies in this model.
More broadly, our results support existing theory of local adaptation but extend it by more completely documenting the genealogical history of adaptive alleles in a metapopulation.

@@

The ancestral marine form of threespine stickleback fish (\textit{Gasterosteus aculeatus}) \revpoint{2}{1}
has given rise to hundreds of thousands to millions of independently derived freshwater populations \revpoint{2}{3}
in recently de-glaciated regions around the Northern Hemisphere 
\citep{Bell_1994, Thompson_1997, Cresko_2007, Hunt_2008} \revpoint{3}{9}
This model organism has provided some of the earliest data showing the heterogeneous nature of divergence across genomes \revpoint{2}{4}
and the much more extensive use of standing genetic variation than once thought 
\citep{schluter2009genetics, DeFaveri_2011, DeFaveri_2013, Roesti_2014, Nelson_2018, Bassham_2018, Nelson_2019, Hohenlohe_2010, Terekhanova_2014, Marques_2016}.  
While geographic isolation often prevents direct migration between freshwater populations, threespine stickleback in them \revpoint{2}{5}
frequently evolve similar phenotypes \citep{Cresko_2004, Colosimo_2004}. 
The most recent evolutionary genomic studies on stickleback document that while the overall dynamic of local adaptation to marine and freshwater habitats has been occurring for millions of years 
\citep{Nelson_2018}, independent local adaptation of marine individuals to freshwater environments has been observed to take place in just tens of generations \citep{Terekhanova_2014, Lescak2015evolution, Bassham_2018}. 
For example, in 1964 the Great Alaskan Earthquake caused an uplift of many islands and coastal regions throughout the Gulf of Alaska. 
Studies of stickleback populations on uplifted Middleton Island showed that newly created freshwater ponds were invaded by the surrounding marine population of stickleback which evolved the freshwater syndrome of phenotypes in less than 50 years 
\citep{Lescak2015evolution, Bassham_2018}. 
Amazingly, the portions of the genomes of these populations that showed increased divergence from the oceanic population
mirrored those previously found to differ between ocean and freshwater populations that
that have been geographically separated for thousands of years 
\citep{Hohenlohe_2010, Bassham_2018}.

@@

To observe the impact of gene flow on selection in the new lakes, we varied the ocean--lake migration rate, $m$, across separate simulations from $5 \times 10^{-5}$ to $5 \times 10^{-1}$. 
Below, we often refer to these migration rates in terms of the number of migrants per lake, per generation, which we denote $M$. Since each lake contains 200 individuals, $M=200m$. 
Many aspects of adaptation changed substantially across this range, including the speed of adaptation, degree of sharing of adaptive alleles between lakes, and the population genetic signals left behind. 
At very low rates of gene flow, each new lake's population adapted almost completely independently through \emph{de novo} mutation, which took a very long time ($\approx 20,000$ generations). 
At very high rates of gene flow, local adaptation was constrained by the large influx of locally maladaptive alleles. % \citep{bolnick2007natural}. 
Between these two extremes, genetic variation that allowed adaptation to freshwater habitats could move relatively easily between lakes. 
Perhaps surprisingly, only a few migrants per generation from the lakes to the ocean were needed to maintain sufficient genetic variation in the ocean to dramatically accelerate adaptation in new lakes.

@@

\subsection*{Widespread allele sharing between lakes}

@@

The numbers in \autoref{fig:MPFAI} strongly suggest that the dramatic increase in speed of local adaptation we observed above occurs because higher gene flow allows sharing of freshwater alleles among populations.
We confirmed this by using recorded tree sequences to identify the origin of each trait-affecting allele in each individual in the new lakes, as defined in the Methods. 
\autoref{fig:Origin} shows that at the lowest rate of gene flow the majority of adaptive alleles are derived from \emph{``de novo''} mutation. 
As gene flow increases, a larger fraction of adaptive alleles derive from pre-existing variation in the marine population at the time of introduction. 
In other words, greater mixing at higher migration rates allows lakes to share alleles instead of developing their own genetic basis of adaptation.

@@

We have thus far found that the speed of adaptation depends strongly on the degree to which alleles can be shared between populations. 
However, the \textit{origin} of the alleles underlying the phenotype is still unknown. 
What is the nature of selection on standing genetic variants? 
In our simulations migration from marine individuals into the new lakes continues throughout introduction and adaptation: but is consistent influx of standing genetic variation a necessity for rapid adaptation of the population?
To answer this, we traced the genealogy of \textit{all} adaptive alleles in \textit{all} individuals from the introduced lakes, after local adaptation.
Surprisingly, for all cases with the exception of $M = 0.01$, we found that the majority of adaptive allele origins traced back the original generation of inhabitants in the lake (\autoref{fig:Origin}). \revpoint{1}{3}
In our simulations,
any reasonable subset of the ancestral population has the potential for rapid adaptation without the need post-colonization hybridization events.

@@

A possible explanation is that at lower rates of gene flow, freshwater alleles present as standing variation in the marine habitat are more tightly linked to marine alleles.  \revpoint{AE}{1}
This implies that adaptation in the new lakes must wait for recombination to separate freshwater and marine adapted alleles.
A freshwater and a marine allele in close proximity
could form an essentially neutral locus,
which would persist in the marine population much longer than a freshwater allele
not ``masked'' in this way.
This suggests that a reasonable portion of freshwater alleles in the marine population
would be linked to marine alleles.
This genetic variation would be more difficult to access, however,
so if there is sufficient additive genetic variation available to a new population in unlinked alleles,
this will be used first.
However, if the genetic variation initially available is not sufficient for a new population to adapt,
then adaptation might need to wait for recombination to separate these haplotypes.
Similar effects have been noted before in simulation \citep[e.g.,][]{yeaman2011genetic}.

@@

In this paper, we have analyzed a relatively realistic simulation study of a coastal meta-population to 
understand how repeated local adaptation operates on complex standing variation across a wide range of gene flow rates.
We have shown that rapid and parallel adaptation similar to that documented in natural stickleback populations
occurs over a realistic range of parameter values.
These simulations therefore provide an in-depth look at (and quantitative proof-of-concept of)
the ``transporter'' hypothesis suggested by \citet{schluter2009genetics}.
Selection is able to rebuild the freshwater haplotype at a rapid pace (in tens of generations) from variation present at migration-selection balance in marine populations 
A finding of practical consequence is that the efficacy of $F_{ST}$-based genomic scans varies significantly with migration rate.
Perhaps surprisingly, we found that the majority of adaptive variant can be traced back to the original generation of new habitats,
with little role for post-colonization migrant alleles. 
In other words, a randomly chosen subset of marine stickleback carry the capacity to rapidly adapt in freshwater habitats without \emph{any} continued connection to the rest of of the species.
Thus, the large marine population was able to harbor and distribute alleles that were deleterious in the ocean habitat, but adaptive in freshwater habitats, to surrounding freshwater populations with only a few migrants per population, per generation. 
% Our simulation findings significantly extend previous empirical and theoretical population genomic studies of stickleback to make testable predictions, and provide a framework for similar studies in other organisms to determine the generality of our findings.

>>>>

Threespine stickleback populations provide a striking example of local adaptation to divergent habitats 
in populations that are connected by recurrent gene flow. 
These small fish occur in marine and freshwater habitats throughout the Northern Hemisphere, and in numerous cases the smaller freshwater populations have been established ``de novo'' from marine colonists. 
Independently evolved freshwater populations exhibit similar phenotypes that have been shown to derive largely from the same standing genetic variants.
Geographic isolation prevents direct migration between the freshwater populations, strongly suggesting that these shared locally adaptive alleles are transported through the marine population.
However it is still largely unknown how gene flow, recombination, and selection jointly impact the standing variation that might fuel this adaptation.
Here we use individual-based, spatially explicit simulations to determine the levels of gene flow that best match observed patterns of allele sharing among habitats in stickleback.
We aim to better understand how gene flow and local adaptation in large metapopulations determine the speed of adaptation and re-use of standing genetic variation. 
In our simulations we find that repeated adaptation uses a shared set of alleles that are maintained at low frequency by migration-selection balance in oceanic populations. 
This process occurs over a realistic range of intermediate levels of gene flow that match previous empirical population genomic studies in stickleback. 
Examining these simulations more deeply reveals how lower levels of gene flow leads to slow, independent adaptation to different habitats, whereas higher levels of gene flow leads to significant mutation load -- but an increased probability of successful population genomic scans for locally adapted alleles. 
Surprisingly, we find that the genealogical origins of most freshwater adapted alleles can be traced back to the original generation of marine individuals that colonized the lakes, as opposed to subsequent migrants. 
These simulations provide deeper context for existing studies of stickleback evolutionary genomics, and guidance for future empirical studies in this model.
More broadly, our results support existing theory of local adaptation but extend it by more completely documenting the genealogical history of adaptive alleles in a metapopulation.

@@

The ancestral marine form of threespine stickleback fish (\textit{Gasterosteus aculeatus}) \revpoint{2}{1}
has given rise to hundreds of thousands to millions of independently derived freshwater populations \revpoint{2}{3}
in recently de-glaciated regions around the Northern Hemisphere 
\citep{Bell_1994, Thompson_1997, Cresko_2007, Hunt_2008} \revpoint{3}{9}
This model organism has provided some of the earliest data showing the heterogeneous nature of divergence across genomes \revpoint{2}{4}
and the much more extensive use of standing genetic variation than once thought 
\citep{schluter2009genetics, DeFaveri_2011, DeFaveri_2013, Roesti_2014, Nelson_2018, Bassham_2018, Nelson_2019, Hohenlohe_2010, Terekhanova_2014, Marques_2016}.  
While geographic isolation often prevents direct migration between freshwater populations, threespine stickleback in them \revpoint{2}{5}
frequently evolve similar phenotypes \citep{Cresko_2004, Colosimo_2004}. 
The most recent evolutionary genomic studies on stickleback document that while the overall dynamic of local adaptation to marine and freshwater habitats has been occurring for millions of years 
\citep{Nelson_2018, Nelson_2019}, independent local adaptation of marine individuals to freshwater environments has been observed to take place in just tens of generations \citep{Terekhanova_2014, Lescak2015evolution, Bassham_2018}. 
For example, in 1964 the Great Alaskan Earthquake caused an uplift of many islands and coastal regions throughout the Gulf of Alaska. 
Studies of stickleback populations on uplifted Middleton Island showed that newly created freshwater ponds were invaded by the surrounding marine population of stickleback which evolved the freshwater syndrome of phenotypes in less than 50 years 
\citep{Lescak2015evolution, Bassham_2018}. 
Amazingly, the portions of the genomes of these populations that showed increased divergence from the oceanic population
mirrored those previously found to differ between ocean and freshwater populations that
that have been geographically separated for thousands of years 
\citep{Hohenlohe_2010, Bassham_2018}.

@@

To observe the impact of gene flow on selection in the new lakes, we varied the ocean--lake migration rate, $m$, across separate simulations from $5 \times 10^{-5}$ to $5 \times 10^{-1}$. 
Below, we often refer to these migration rates in terms of the number of migrants per lake, per generation, which we denote $M$. Since each lake contains 200 individuals, $M=200m$. 
Many aspects of adaptation changed substantially across this range, including the speed of adaptation, degree of sharing of adaptive alleles among lakes, and the population genetic signals left behind. 
At very low rates of gene flow, each new lake's population adapted almost completely independently through \emph{de novo} mutation, which took a very long time ($\approx 20,000$ generations). 
At very high rates of gene flow, local adaptation was constrained by the large influx of locally maladaptive alleles. % \citep{bolnick2007natural}. 
Between these two extremes, genetic variation that allowed adaptation to freshwater habitats could move relatively easily between lakes. 
Perhaps surprisingly, only a few migrants per generation from the lakes to the ocean were needed to maintain sufficient genetic variation in the ocean to dramatically accelerate adaptation in new lakes.

@@

\subsection*{Widespread allele sharing among lakes}

@@

The numbers in \autoref{fig:MPFAI} strongly suggest that the dramatic increase in speed of local adaptation we observed above occurs because higher gene flow allows sharing of freshwater alleles among populations.
We confirmed this finding by using recorded tree sequences to identify the origin of each trait-affecting allele in each individual in the new lakes, as defined in the Methods. 
\autoref{fig:Origin} shows that at the lowest rate of gene flow the majority of adaptive alleles are derived from \emph{``de novo''} mutation. 
As gene flow increases, a larger fraction of adaptive alleles derive from pre-existing variation in the marine population at the time of introduction. 
In other words, greater mixing at higher migration rates allows lakes to share alleles instead of developing their own genetic basis of adaptation.

@@

We have thus far found that the speed of adaptation depends strongly on the degree to which alleles can be shared between populations. 
However, the \textit{origin} of the alleles underlying the phenotype is still unknown. 
In our simulations migration from marine individuals into the new lakes continues throughout introduction and adaptation: but is consistent influx of standing genetic variation a necessity for rapid adaptation of the population?
To answer this, we traced the genealogy of \textit{all} adaptive alleles in \textit{all} individuals from the introduced lakes, after local adaptation.
Surprisingly, for all cases with the exception of $M = 0.01$, we found that the majority of adaptive allele origins traced back to the original generation of inhabitants in the lake (\autoref{fig:Origin}). \revpoint{1}{3}
In our simulations,
any reasonable subset of the ancestral population has the potential for rapid adaptation without the need post-colonization hybridization events.

@@

A possible explanation is that at lower rates of gene flow, freshwater alleles present as standing variation in the marine habitat are more tightly linked to marine alleles.  \revpoint{AE}{1}
This implies that adaptation in the new lakes must wait for recombination to separate freshwater and marine adapted alleles.
A freshwater and a marine allele (of roughly equal and opposite effects) in close proximity
could form an essentially neutral locus,
which would persist in the marine population much longer than a freshwater allele
not ``masked'' in this way.
This suggests that a reasonable portion of freshwater alleles in the marine population
would be linked to marine alleles.
This genetic variation would be more difficult to access, however,
so if there is sufficient additive genetic variation available to a new population in unlinked alleles,
this will be used first.
However, if the genetic variation initially available is not sufficient for a new population to adapt,
then adaptation might need to wait for recombination to separate these haplotypes.
Similar effects have been noted before in simulation \citep[e.g.,][]{yeaman2011genetic}.

@@

In this paper, we have analyzed a relatively realistic simulation study of a coastal meta-population to 
understand how repeated local adaptation operates on complex standing variation across a wide range of gene flow rates.
We have shown that rapid and parallel adaptation similar to that documented in natural stickleback populations
occurs over a realistic range of parameter values.
These simulations therefore provide an in-depth look at (and quantitative proof-of-concept of)
the ``transporter'' hypothesis suggested by \citet{schluter2009genetics}.
Selection is able to rebuild the freshwater haplotype at a rapid pace (in tens of generations) from variation present at migration-selection balance in marine populations 
A finding of practical consequence is that the efficacy of $F_{ST}$-based genomic scans varies significantly with migration rate.
Perhaps surprisingly, we found that the majority of adaptive variant can be traced back to the original generation of new habitats,
with little role for post-colonization migrant alleles. 
In other words, a randomly chosen subset of marine stickleback carry the capacity to rapidly adapt in freshwater habitats without \emph{any} continued connection to the rest of of the species.
Thus, the large marine population was able to harbor and distribute alleles that were deleterious in the ocean habitat, but adaptive in freshwater habitats, to surrounding freshwater populations with only a few migrants per population, per generation. A few good stickleback suffice.
% Our simulation findings significantly extend previous empirical and theoretical population genomic studies of stickleback to make testable predictions, and provide a framework for similar studies in other organisms to determine the generality of our findings.
